Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model

### 2023-05-16 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
ELGM for #40

@@

>>>>

<!-- https://github.com/athowes/elgm-inf/blob/fd4b0bb6b9ba93ac277de7aad47a94581e8fc900/src/docs_18-04-23-explainer/18-04-23-explainer.Rmd -->

@@

# Latent Gaussian models

* In an LGM, the conditional mean depends on exactly one structured additive predictor
\begin{align*}
y_i &\sim p(y_i \, | \, \eta_i, \theta_1), \quad i \in [n]\\
\mu_i &= \mathbb{E}(y_i \, | \, \eta_i) = g(\eta_i), \\
\eta_i &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}),
\end{align*}
where $\beta_0$, $\{\beta_j\}$ and $\{f_k(\cdot)\}$ have Gaussian priors (and can be collected into the latent field $x$)

# Extended latent Gaussian models

* ELGM remove this requirement such that
$$
\mu_i = g_i(\eta_{\mathcal{J}_i})
$$
where $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$ and $\mathcal{J}_i$ is some set of indices
* Let $\dim(\eta) = N_n$ with $\mathcal{J}_i \subseteq \{1, \ldots, N_n\}$
  * $N_n < n$: more data points than structured additive predictors
  * $N_n = n$: as many data points as structured additive predictors (LGM case)
  * $N_n > n$: fewer data points than structured additive predictors
* The $g_i$ allow for a higher degree of non-linearity in the model

# Why is Naomi an ELGM?

| | Things that make Naomi an ELGM |
|-|---------------|
|1.| Incidence depends on adult prevalence and coverage |
|2.| Incidence linked non-linearly to recent infection |
|3.| ANC offset from household survey |
|4.| ART coverage and recent infection are products |
|5.| Aggregation of finer processes |
|6.| ART attendance uses a multinomial |
|7.| Multiple link functions |

# 1. Incidence depends on adult prevalence and coverage

* Linear predictor for incidence contains aggregated prevalence and coverage
$$
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda.
$$
* Here $\log(\rho_{x}^{\text{15-49}})$ and $\log(1 - \omega \cdot \alpha_{x}^{\text{15-49}})$ are not going to be Gaussian

# 2. Incidence and prevalence linked to recent infection

$$
\kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right)
$$

* Here $\Omega_T$ and $\beta_T$ are Gaussian and have strong priors depending on the particular survey

# 3. ANC offset from household survey

* Linear predictors for ANC indicators contain nested in them the linear predictors for household survey indicators
\begin{align*}
\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}}. 
\end{align*}
* Here $\text{logit}(\rho_{x, F, a})$ and $\text{logit}(\alpha_{x, F, a})$ *are* Gaussian, but we have dependency of $\mu_i$ on two $\eta_i$

# 3. ANC offset from household survey

* Note that `R-INLA` does have the `copy` feature $\eta^\star = A \eta$ where $A$ is $n \times n$^[I've also seen it claimed that it could be $m \times n$ where $m \neq n$, so I'm unsure which is right]
$$
\begin{pmatrix}
\eta_1^\star \\
\eta_2^\star
\end{pmatrix}
=
\begin{pmatrix}
1 & 1 \\
0 & 1
\end{pmatrix}
\begin{pmatrix}
\eta_1 \\
\eta_2
\end{pmatrix}
=
\begin{pmatrix}
\eta_1 + \eta_2 \\
\eta_2
\end{pmatrix}
$$
* By having effects only apply to a subset of indices (`idx = c(NA, 1)` say) perhaps it can work

Warning!

\begin{enumerate}
    \item Is an LGM $\nRightarrow$ can be fit with \texttt{R-INLA}
    \item Can be fit with \texttt{R-INLA} $\nRightarrow$ is an LGM
\end{enumerate}

# 4. ART coverage and recent infection are products

* In the household survey, say, individuals who are taking ART or have been recently infected must be HIV positive
\begin{align*}
y^{\hat \alpha}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \alpha_{x, s, a}), \\
y^{\hat \kappa}_{x, s, a} &\sim \text{xBin}(m_{x, s, a}, \rho_{x, s, a} \cdot \kappa_{x, s, a}).
\end{align*}
* $\text{logit}(\rho_{x, s, a})$ and $\text{logit}(\alpha_{x, s, a})$ are Gaussian, but we're taking a product here
* $\kappa_{x, s, a}$ is more complicated: a function of incidence, prevalence, mean duration of recent infection and false recent ratio

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
Warning! These equations as writen do not appear in Naomi: instead there are aggregated versions, more about this soon.
\end{tcolorbox}
\end{center}

# 5. Aggregation of finer processes

* There are many instances of models being placed on aggregate quantities
\begin{align*}
y^{\hat \theta}_{\mathcal{I}} &\sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}}), \\
\rho_{\mathcal{I}} &= \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}.
\end{align*}
* Here we have $|\mathcal{I}|$ linear predictors being informed by one observation
* Known as disaggregation regression

# 6. ART attendance uses the multinomial

# 7. Multiple link functions

* The Naomi model uses both $\text{logit}$ and $\log$ (inverse) link functions (not even considering the constructed quasi-link functions)
* For LGMs there is only one $g$, whereas ELGMs allow $g_i$
* In `R-INLA` it is possible for `y` to be a matrix where each column contains observations with shared likelihood family (and hyperparameters) and `family = c("family1", "family2", ...)`

# Statistical software

|           | Interface `y ~ 1 + x`  | General |
|-----------|------------|---------|
| Example   | `brms`     | Stan    |
| Users     | Scientists | Statisticians |
| Benefits  | Ease of use, sensible defaults, trust | Flexibility, development |
| Drawbacks | Being fenced in | Barrier to entry, failing silently |

\begin{center}
\begin{tcolorbox}[width=0.9\textwidth, colframe={title}, colback={white}, title={}]
\texttt{athowes/multi-agyw} case-study: attempting to define $\text{AR}(1) \times \text{Besag} \times \text{IID}$ random effects in \texttt{R-INLA}
\end{tcolorbox}
\end{center}
